---
title: 'A3: Incarceration'
output:
  html_document: default
  pdf_document: default
---

### Introduction

The problem domain of this dataset that I chose to study and analyze
is the disproportionate makeup of prisons in the United States
regarding the population and ratio of Black inmates to White inmates.
I want to analyze the variations in ratios of Black to White inmates
across prisons specifically, since this assignment is related to
incarceration, so I'm only using the CSV file related to prison
population per county/state. My variables of interest in this data set
are year, state, total_prison_pop, black_prison_pop, and
white_prison_pop. I specifically chose to only analyze the Black and
White populations because Black people make up a disproportional amount
in U.S prisons compared to White people, so observing how those ratios
vary across states would help corroborate the findings showcased in the documentary 13th, while heavily focused on this unbalanced population
ratio. 


### Summary Information

Write a summary paragraph of findings that includes the 5 values calculated from your summary information R script


My two main variables of interest are Black to White prison population
ratios and total average population counts. To further explore these
variables and get myself familiar with them, I first calculated the total_average_black_population and total_average_white_population,
which was about 247 for White and 162 for Black. These values are not
a good picture of the real proportions as this data spans across almost
50 years worth of data, most of which are NA or extremely low numbers.
Then I calculated the average black_to_white_ratio, grouped by state to
create a data frame of only the states and their respective
black_to_white_ratio. From this data frame, called average_prison_ratios,
I arranged it in descending order from greatest ratio to lowest ratio
and filtered out the top 10 states into a variable called
sorted_ratio_states. The highest black_to_white_ratio among these states
was "LA" or Louisiana, stored in the variable highest_state. The black_to_white_ratio for LA was around 3.165, meaning there were 
on average over 3 times as many Black inmates than White inmates in
Louisiana prisons, stored in the variable highest_ratio. This is a
significant number which will be studied further on, along with the other states, later on in this analysis. 

### The Dataset

Who collected the data?

This data was collected by the Vera Institute of Justice.

How was the data collected or generated? 

The data was collected through an assembly of information collected by
the U.S Department of Justice Statistics (BJS), with additional data
from state departments of correction. 

Why was the data collected?  
The data was collected to understand the contours and patterns of
incarceration in the U.S, especially in context of race, as much of the
information on prison populations have been insufficient to have more
robust and theoretically-grounded studies. The county-level variability
within these data sets allow for a deeper analysis. 

How many observations (rows) are in your data?

There are 153611 observations in the prison dataset. 

How many features (columns) are in the data? 

There are 38 columns/variables in the prison dataset. 

What, if any, ethical questions or questions of power do you need to consider when working with this data? 

There are some important ethical questions that must be asked when 
working and analyzing this data, mostly related to the lack of context surrounding these observations and population counts of prisons.
We don't know if the data about certain population sizes in prisons,
especially in regards to race, is even accurate or represented properly.
We also must ask ourselves, as students, the kinds of conclusions we will
be drawing from these types of data, which is solely based on population
counts, and gives no insight to the treatment, crimes, or situations of
the people who make up this data. These data sets are more than just
numbers, they represent actual human beings. 


What are possible limitations or problems with this data?   (at least 200 words)

There are a lot of limitations with this data, the biggest one being
the massive amounts of missing population data. It was shocking to see
that numerous states contained absolutely zero information about their
prison populations and racial/gender makeup over the span of nearly 50
years for some regions. This creates a lot of problems in the analysis
and interpretation of the graphs, tables, and findings since a large
chunk of the data that could either support or refute a certain conclusion
is missing. My variable of interest, the ratio of Black to White inmate populations is thus severely unsupported due to this missing data.
It raises questions as to why these states either, refused to report
or didn't report, on their prison population counts. Another limitation
to the data, that I mentioned about when talking about ethics, was the
lack of context given to these counts. While these data sets are
incredible on their own and contain a mass amount of prison data that
can help draw out importation patterns in our country's incarceration
history, the humanity and stories behind these people are lost. What's
to say a prison contained a small-amount of Black inmates; the immediate conclusion would be that this prison was not disproportionately 
targeting people of color, yet perhaps those small amount of Black
inmates were being treated ten times more poorly than their white
counterparts or were locked up for crimes they didn't commit? So much
context is lost through this data and is why the conclusions that are
made from this analysis should not be absorbed as the cold and hard
truth about our prison system; as most of the issues stay hidden and are systemically embedded into institutions more powerful than prisons. 


### Trends Over Time Chart

``````{r, echo = FALSE, code = readLines("chart1_example.R")}
```


The first chart that I made was a line graph displaying the trend of black_to_white_ratio for the top 10 states in the U.S overtime. It ranges
from 1970 to 2018. As we can see from the graph, there is a clear upwards
trend in the black_to_white_ratio for all 10 states, meaning that the disproportionate makeup of Black inmates in their prisons was rapidly increasing, some jumping from 1 to 4 times as many Black inmates than
White, most noticeably for NJ. We can also see where important political
events and laws were passed in regards to incarceration, the most apparent
one being the massive spike in the mid to late 90's, when President Clinton created the infamous crime bill and ultimately doubled the amount of Black inmates in prisons across the U.S. There has been a gradual decrease since
2000, but overall the black_to_white ratio for these states are still too
high to be ignored. 


### Variable Comparison Chart

``````{r, echo = FALSE, code = readLines("chart2_example.R")}
```

This chart compares the correlation between average prison population and average Black to White prison population ratios for all states in this
dataset. I created this scatter plot to see if there was any correlation
between larger, more populated prisons, and their demographic ratios between Black and White inmates. There is a slight upwards trend based on this
graph that may be evident to larger prisons often housing a disproportionate amount of Black inmates, but overall not enough to fully conclude it, as
these averages are often too low due to the time span and the 10 missing
values from states who reported zero information on their populations. 

### Map

``````{r, echo = FALSE, code = readLines("map_example.R")}
```

This is a map showcasing the geographical distribution of the variation in black_to_white_ratios across the United States. The legend on the side represents the color shading that corresponds to the rounded ratios
(1, 2, and 3 meaning that the Black inmate population is x times as large
as the White inmate population). There is more red shading in the
Southern and East Coast states, states where there are larger populations
of Black people, with darker shading in the Central and more Western
states, as well as the grey states indicating completely NA
ratios/populations that were unfortunately present in the data set. 

